---
title: "DS 202 Final Project"
author: "Nathan Catalano and Brett Simmons"
date: "2023-05-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# install.packages("Lahman")
library(dplyr)
library(ggplot2)
library(Lahman)
```

# Primary Research Questions

### How have various statistics and player performance evolved over time?

### Looking from the only the year 2000 and up how have various statistics and player performance evolved over time?

# Relevant Variabes 

```{r}
data(Batting)
data(Pitching)
data(People)
data(Parks)
data(Teams)
```

### League Batting average evolution

The league wide batting average has varied a lot over the seasons. It was highest in the late 1800's and since then has remained pretty consistently between .240 and .270 since the 1940's.

```{r}
avgs = Batting %>%
  group_by(yearID) %>%
  summarize(league_avg = sum(H) / sum(AB))

ggplot(avgs, aes(x = yearID, y = league_avg)) +
  geom_line() +
  labs(x = "Year", y = "League Batting Average", title = "League Batting Average over Time")
```

### Total number of homeruns per season 

The total number of home runs has been rising steadily. It peaked in 2019 at 6,776 with the next closest season being 2017 at 6,105.

```{r}
hr_total = Batting %>%
  group_by(yearID) %>%
  summarize(total_HR = sum(HR))

ggplot(hr_total, aes(x = yearID, y = total_HR)) +
  geom_line() +
  labs(x = "Year", y = "Total Home Runs", title = "Total Home Runs by Season")

```

### Total number of hits per season

Besides the fluctuations in the early seasons, the total number of hits was steady from the 1920's up until the 1960's when it began to increase. The notable outliers are during lockouts/strikes and the shortened covid season. After a slight decline in the early 2000's its beginning to level out again around 42,000 hits per season.

```{r}
hits_total = Batting %>%
  group_by(yearID) %>%
  summarize(total_hits = sum(H))

ggplot(hits_total, aes(x = yearID, y = total_hits)) +
  geom_line() +
  labs(x = "Year", y = "Total Hits", title = "Total Hits by Season")

```

### Total walks per season

Following a very similar trend to total hits by season, the number of walks has steadily been increasing. It peaked in 2000 and is leveling off around 16,000 per year.

```{r}
walks_total = Batting %>%
  group_by(yearID) %>%
  summarize(total_walks = sum(BB))

ggplot(walks_total, aes(x = yearID, y = total_walks)) +
  geom_line() +
  labs(x = "Year", y = "Total Walks", title = "Total Walks by Season")

```

### Total stolen bases per season

In the 1800's for a short while, stolen bases were credited when a base runner reached an extra base on a base hit from another player, hence why steals were so high. After that rule change the total number of stolen bases dropped immensely. During the dead ball era in the 1900's it rose again, peaking in 1914. After declining from then on stealing began to rise in the late 60's and 70's, close to current rates now. 

```{r}
sb_total = Batting %>%
  group_by(yearID) %>%
  summarize(total_SB = sum(SB))

ggplot(sb_total, aes(x = yearID, y = total_SB)) +
  geom_line() +
  labs(x = "Year", y = "Total Stolen Bases", title = "Total Stolen Bases by Season")

```

### League average era per season

The outliers in 1893 and 1894 occurred because the pitching distance was increased. Since then the league wide average era by season has remained relatively consistent with minor fluctuations season to season. 

```{r}
era_avgs = Pitching %>%
  group_by(yearID) %>%
  summarize(avg_ERA = mean(ERA, na.rm = TRUE))

# Create a line chart of average ERA over time
ggplot(era_avgs, aes(x = yearID, y = avg_ERA)) +
  geom_line() +
  labs(x = "Year", y = "Average ERA", title = "Average ERA by Season")

```

### Average height of players by birth year (1850 to 1999)

The average height of players has been steadily increasing over the years. The average height is now slightly below 6'2". 

```{r}
People$birthYear = as.numeric(People$birthYear)
greater_1850 = filter(People, birthYear >= 1850 & birthYear < 1999)

avg_height = greater_1850 %>%
  group_by(birthYear) %>%
  summarize(mean_height = mean(height, na.rm = TRUE))

ggplot(avg_height, aes(x = birthYear, y = mean_height)) +
  geom_line() +
  labs(x = "Birth Year", y = "Average Height (inches)", title = "Average Height by Birth Year (1850 - Present)")

```

### Average weight of players by season 

As expected, the average weight of players has also steadily been increasing. The most recent average weight was in 2021 at 211 lbs. 
```{r}
People_filtered = filter(People, !is.na(height) & !is.na(weight))
Batting_filtered = select(Batting, playerID, yearID) %>% distinct()

joined = left_join(People_filtered, Batting_filtered, by = "playerID")

avg_weight = joined %>%
  group_by(yearID) %>%
  summarize(mean_weight = mean(weight))

ggplot(avg_weight, aes(x = yearID, y = mean_weight)) +
  geom_line() +
  labs(x = "Year", y = "Average Weight (lbs)", title = "Average Weight of MLB Players by Season")

```

###Which players have gotten the most hits since 2000?

```{r}
yearhits = Batting %>% filter(yearID >= 2000) %>% group_by(playerID) %>% summarise( total.hits= sum(H)) %>% arrange(desc(total.hits))
```

```{r}
yearhitstop=yearhits[1:10, ]
ggplot(yearhitstop, aes(x=playerID, fill=playerID))+geom_bar(aes(weight=total.hits))
```
The top 10 players with the most hits since 2000 are: Albert Pujols, Ichiro Suzuki, Miguel Cabrera, Adrian Beltre, Derek Jeter, Robinson Cano, Carlos Beltran, Jimmy Rollins, Nick Markakis, and David Ortiz.


###Which team has scored the most runs since 1998?

The reason that 1998 is chosen is because that is when the league introduced the Arizona Diamondbacks and the Tampa Bay Devil Rays which gave the league 30 teams which still stands today

```{r}
run= Batting %>% filter(yearID>=1998) %>% group_by(teamID,yearID) %>% summarise(total.runs=sum(R))

high = run %>% group_by(yearID) %>% slice_max(total.runs)

```

```{r}
ggplot(high, aes(x=yearID, y=total.runs,color=teamID))+geom_point()
```
The chart shows the league leaders in runs scored for each year from 1998 to 2021. For example, the New York Yankees led the league in 1998 and the Houston Astros led the league in 2021.  

### Are the number of international players increasing?

```{r}
Country= joined %>% filter(yearID>=1980,birthCountry !="USA") %>% group_by(yearID) %>% summarise(total.players=n())
```

```{r}
Country %>% ggplot(aes(x=yearID,y=total.players))+geom_line()
```

The number of international players has about quadrupled from the year 1980 to 2021


### What differences are there in ERA for pitchers in the 1980s and 1990s compared to since 2000s

```{r}
era80 = Pitching %>% filter(between(yearID,1980,1999)) %>% filter(G > 20)
era00 = Pitching %>% filter(yearID >=2000) %>% filter(G>20)
```

```{r}
ggplot(era80, (aes(x = ERA))) + geom_histogram()+geom_vline(aes(xintercept = mean(ERA)),col='red',size=2)
```

```{r}
ggplot(era00, (aes(x = ERA))) + geom_histogram()+geom_vline(aes(xintercept = mean(ERA)),col='red',size=2)
```

Average pitchers' ERA is higher since 2000 compared to the 1980s and 1990s

### How are the debut ages of players changing?
```{r}
joined$debut=as.Date(joined$debut)
library("lubridate")
joined$debut.age=time_length(difftime(joined$debut,joined$birthDate),"years")
```

```{r}
joined2 = joined %>% mutate(debut=ymd(debut)) %>% mutate_at(vars(debut),funs(year,month,day))
```

```{r}
age = joined2 %>% filter(year>=1980) %>% group_by(year) %>% summarise(debut.age)
```

```{r}
ggplot(age, (aes(x=year, y=debut.age)))+geom_boxplot()+coord_flip()+facet_wrap(~year)
```
